Increasing productivity amid stable nutrient regmines in Rhode Island Lakes and Reservoirs

### 2019-11-07 nalms_2019_talk/increasing_chla_stable_nutrients_talk.Rmd
progress on slides!

## Nutrients and Productivity
::::::::::::::{.columns}
:::{.column}
- Nutrients drive growth
- One goal is to reduce increased phyto
- Long-term data needed
:::
:::{.column}

@@

![](figures/tn_tp_chla_plot.jpg)
:::
::::::::::::::

## Nutrient Management

- What has happened with P and N reductions?
- What 
- Did it work?

@@

content

@@

- VT example
- Other examples?
- What about RI?

## Watershed Watch

- Details on WW program and data collected

## Watershed Watch: filtered for trend analysis

- What we filtered
- resultant datasets details

>>>>

## What drives changes in lake productivity?
::::::::::::::{.columns}
:::{.column}
- Nutrients
- Temperature
- Other things
:::
:::{.column}

@@

![](figures/tn_tp_chla_plot.jpg)

![](figures/temp_chla_plot.jpg)
:::
::::::::::::::

## The management problem

- Desire to reduce and manage nutrient pollution
- Phosphorus Bans
- Nutrient Criteria
- Enacted piecemeal over last ~20 years
- Do the data show any change?

@@

- Oliver et al (2017) Unexpected stasis in a changing world: Lake nutrient and
chlorophyll trends since 1990. Global Change Biology. DOI: 10.1111/gcb.13810

![](figures/10.1111-gcb.13810Figure138100001.jpg)

@@

- Mathews et al (2018) Is Vermont losing its oligotrophic lakes?. LakeLine. 38-2.
- Rhode Island?

![](figures/mathews_vt_olgio_lakeline_fig2.jpg)

## URI Watershed Watch

- Volunteer monitoring program
  - ~400 volunteers
- Started in 1988 with 14 lakes
- Now
  - 250+ sites
  - 120+ waterbodies
- Rigorous QA/QC
- Data used by RI DEM and US EPA

## URI Watersehd Watch: Lake Sampling

- May to October
- Weekly
  - Water Clarity
  - Temperature
- Every Other Week
  - Chlorophyll
  - Dissolved Oxygen
- 3 Times per Season
  - Nutrients
  - Alakalinity
  - pH
  - Bacteria

## URI Watershed Watch: Filtered for Trend Analysis

- Parameters: Total Nitrogen, Total Phosphorus, N:P, Temperature, Chlorophyll
- Years: 1993 to 2016
- Months: May through October
- Depths: <= 2 meters
- Sites: At least 10 years of data
- Total of 69 Sites

![](figures/ww_map.jpg)
